
## KeynessMasures

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

An R package including multiple measures of word keyness used in computational
linguistics. Introducing both statistical significance and effect size
measures in several simple functions.


This package currently supports the following measures:

1.  **Statistical significance:** Log-likelihood ratio, Bayesian Information Criterion.

2.  **Effect size:** Effect Size of Log Likelihood, %DIFF, The Relative Risk,
The Log Ratio measure, The Odds Ratio.

For more details on using the effect size measures, consult the vignette 
"KeynessMeasures: Introduction to effect size measures". 

## Installing the package

This package can be installed from GitHub, using devtools. 

To download the package from GitHub, use the following command:
```{r, eval=F, message=FALSE, warning=FALSE}
devtools::install_github("amacanovic/KeynessMeasures")
```

Then, load the package:
```{r message=FALSE, warning=FALSE}
library(KeynessMeasures)
```

For more information on included measures, type:
```{r}
vignette("Effect_size_measures")
```

For a detailed tutorial vignette, type:
```{r}
vignette("Keyness_Measures_Tutorial")
```


## Demonstration

Using the `keyness_measure_calculator()` function, one can easily obtain
several measures of word keyness in a target corpus compared to the 
reference corpus.

Demonstration using the Jane Austen data from the **janeaustenr** package (Silge, 2017).
We will be exploring key words in her novel "Emma" compared to 5 other novels.
```{r, warning=FALSE, message=FALSE}

jane_austen_data <- janeaustenr::austen_books()

```

First, obtaining the word frequencies in "Emma" (target corpus) and the other
novels (reference corpus) using the `frequency_table()` function:
```{r}
frequency_table <- frequency_table_creator(jane_austen_data,
                                           text_field = "text",
                                           grouping_variable = "book",
                                           grouping_variable_target = "Emma",
                                           lemmatize = TRUE,
                                           remove_punct = TRUE,
                                           remove_symbols = TRUE,
                                           remove_numbers = TRUE,
                                           remove_url = TRUE)
```

Then, calculating all the keyness measures using the `keyness_measure_calculator()` function,
sorting the entries by highest values of log likelihood:

```{r}
keyness_measures <- keyness_measure_calculator(
  frequency_table,
  log_likelihood = TRUE,
  ell = TRUE,
  bic = TRUE,
  perc_diff = TRUE,
  relative_risk = TRUE,
  log_ratio = TRUE,
  odds_ratio = TRUE,
  sort = "decreasing",
  sort_by = "log_likelihood")
```

Displaying the first 10 rows of results sorted by log likelihood values:
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

keyness_measures[1:10,] %>%
  kbl() %>%
  kable_classic(full_width = F, html_font = "Cambria")
```
Instead, sorting the words by %DIFF measure:

```{r, echo=FALSE, message=FALSE, warning=FALSE}

keyness_measures_2 <- keyness_measure_calculator(
  frequency_table,
  log_likelihood = TRUE,
  ell = TRUE,
  bic = TRUE,
  perc_diff = TRUE,
  relative_risk = TRUE,
  log_ratio = TRUE,
  odds_ratio = TRUE,
  sort = "decreasing",
  sort_by = "perc_diff")

keyness_measures_2[1:10,] %>%
  kbl() %>%
  kable_classic(full_width = F, html_font = "Cambria")
```


